# Exploratory Data Analysis EDA

## 1.1 Numerical Summaries
The summary descriptive characteristics of a sample of objects, that is, a subset of the population, are called statistics. Sample statistics can have different values, depending on how the sample of the population was chosen. Statistics are denoted by various symbols, but (almost) never by Greek letters e.g. sample mean, $\bar y$ and sample standard deviation, $s$.

### 1.1.1 Measures of Central Tendency

These statistics are also sometimes referred to as measures of location.

**ARITHMETIC MEAN**

The most widely used measure of central tendency is the arithmetic mean or average. The population mean, $\mu$ (“mu”) is the sum of all the values of the variable under study divided by the total number of objects in the population, Each value is algebraically denoted by a $y$ with
a subscript denotation $i$. E.g. a small theoretical population whose objects had values 1, 6, 4, 5, 6, 3, 8, 7 would be denoted $y_1 = 1, y_2 = 6, y_3 = 4, y_4 = 5, y_5 = 6, y_6 = 3, y_7 = 8, y_8 = 7$.

**Noe**. Some texts will use $X$ instead of $x$ or $Y$ instead of $y$ as the symbol for a value.

We would denote the population size with a capital $N$. In our theoretical population $N = 8$.

The **population mean** $\mu$ would be

$\frac{1+6+4+5+6+3+8+7}{8}=5$

The algebraic shorthand formula for a population mean is

$\mu = \frac{1}{N}\sum_{i=1}^{N} y_i$

The Greek letter $\Sigma$ (“sigma”) indicates summation, the subscript $i = 1$ means to start with the first observation, and the superscript $N$ means to continue until and including the $N$th observation.

For the example above,

$\sum_{i=2}^{5} y_i=y_2+y_3+y_4+y_5=6+4+5+6=21$

To reduce the clutter, if the summation sign is not indexed, for example $y_i$, it is implied that the operation of addition begins with the first observation and continues through the last observation in a population, that is,

$\sum_{i=1}^{N} y_i = \sum y_i$

The **sample mean** is defined by

$\bar y = \frac{1}{n}\sum_{i=1}^{n} y_i$

where $n$ is the sample size.

The symbol $\bar y$ (read “y bar”) indicates that the observations of a subset of size n from a population have been averaged. $y$ is fundamentally different from $\mu$ because samples from a population can have different values for their sample mean, that is, they can vary from sample to sample within the population. The population mean, however, is constant for a given population.

Again consider the small theoretical population $1, 6, 4, 5, 6, 3, 8, 7$. A sample size of 3 may consist of $5, 3, 4$ with $\bar y =6$ OR it could be $1,3,5$ with $\bar y = 3$

Each sample mean $\bar y$ is an unbiased estimate of $m$ but depends on the values included in the sample and sample size for its actual value. We would expect the average of all possible $y$’s to be equal to the population parameter, $\mu$. This is, in fact, the definition of an unbiased estimator of the population mean.

The R command is `mean()`.

```{r}
y <- c(1, 6, 4, 5, 6, 3, 8, 7)
mean(y)
```


The Excel command is `=AVERAGE()`

**MEDIAN**

The median is the “middle” value of an ordered list of observations. The **population** median $M$ is the $\left( \frac{N+1}{2} \right)th$ sorted value, where $N$ is the population size. Note that this parameter is not a
Greek letter and is seldom computed in practice. A **sample median** $\tilde y$ (read “y tilde”) is the statistic used to approximate or estimate the population median. $\tilde y$ is the $\left( \frac{n+1}{2} \right)th$ sorted value where n is the sample size.

The R command is `median()`.

```{r}
y <- c(1, 6, 4, 5, 6, 3, 8, 7)
median(y)
```

The Excel command is `=MEDIAN()`.

**MODE**

The mode is the most frequently occurring value in a data set.

There is no direct command in R, the following code is an example of how it can be calculated.

```{r}
mode_function <- function(x) {
  uniqx <- unique(x)
  uniqx[which.max(tabulate(match(x, uniqx)))]
}

# Example usage
y <- c(1, 6, 4, 5, 6, 3, 8, 7)
mode <- mode_function(y)
print(mode)
```

The Excel formula is =MODE( … ).

**OVERVIEW OF MEASURES OF CENTRAL TENDENCY:**

- The mean is a purposeful measure only for a quantitative variable, whether it is continuous (e.g. height) or discrete (e.g. number of nematodes).
- The median can be calculated whenever a variable can be ranked (including when the variable is quantitative).
- The mode can be calculated for categorical variables, as well as for quantitative and ranked variables.
- The sample median expresses less information than the sample mean because it utilizes the ranks and not the actual values of each measurement.
- The median, however, is resistant to the effects of outliers. Extreme values or outliers in a sample can drastically affect the sample mean, while having little effect on the median.


## 1.1.2 Measures of Spread

Measures of central tendency alone are not sufficient to fully describe a data set. The following figure illustrates 3 distributions that all have the same mean but different levels of dispersion or spread.

```{r, fig.cap="Figure 1.1 A, B & C are distributions with the same mean (zero) but varying standard deviations."}
library(ggplot2)

# Define the data for the three normal distributions
mean_value <- 0
std_dev_A <- 1   # Least spread
std_dev_B <- 2
std_dev_C <- 3   # Most spread

# Create a data frame for plotting
x_values <- seq(-10, 10, length.out = 300)
normal_data <- data.frame(
  x = c(x_values, x_values, x_values),
  y = c(dnorm(x_values, mean_value, std_dev_A), 
        dnorm(x_values, mean_value, std_dev_B), 
        dnorm(x_values, mean_value, std_dev_C)),
  curve = factor(c(rep("A", length(x_values)), 
                   rep("B", length(x_values)), 
                   rep("C", length(x_values)))
  )
)

# Generate the plot
ggplot(normal_data, aes(x = x, y = y, color = curve)) +
  geom_line() +
  labs(title = "Normal Distributions with Different Spreads",
       x = "Value",
       y = "Density") +
  scale_color_manual(values = c("red", "green", "blue"),
                     labels = c("A: Least Spread", "B", "C: Most Spread")) +
  theme_minimal()
```


In graph A, most of the values are concentrated around the mean. It has less dispersion (or spread of values) than the other distributions. Graph C has more dispersion than the others. Its data is more “spread” out.
A measure of dispersion provides some indication of the amount of variation that the data exhibits.

**RANGE**

The simplest measure of dispersion or “spread” is the range – the difference between the largest and smallest observations in a group of data.

The sample range is a crude and biased estimator of the population range as its dependent on the composition and size of the sample you’ve taken. [It’s unlikely that the sample will include the largest and smallest values from the population, so the sample range usually underestimates the population range and is, therefore, a *biased* estimator.]

The R command is `max(y)-min(y)`

```{r}
y <- c(1, 6, 4, 5, 6, 3, 8, 7)
max(y)-min(y)
```

The Excel command is `=MAX()-MIN()`

**INTERQUARTILE RANGE**

Rather than describe variability in terms of variation around the mean, we more directly quantify the “spread”. Just as the median divides the sample into two, the quartiles divide the sample into four groups:

- 25% of observations $\le$ lower quartile (Q1) 
- 50% of observations $\le$ median (Q2) 
- 75% of observations $\le$ upper quartile (Q3)

The example data below (the number of days pigs take to reach bacon weight) has been sorted from lowest to highest.

98 100 100 103 105 107 110 113 115 

The lower quartile Q1 is the $\left( \frac{n+1}{4} \right)th$ sorted value = $\left( \frac{9+1}{4} \right)th = 2.5th$ sorted value.

This means we need to obtain a *weighted* average of the 2nd and 3rd sorted value: $0.5 \times 100 + 0.5 \times 100 = 100 \text{ days}$

The upper quartile Q3 is the $\left( \frac{3(n+1)}{4} \right)th$ sorted value = $\left( \frac{3(9+1)}{4} \right)th = 7.5th$ sorted value.

This means we need to obtain a *weighted* average of the 2nd and 3rd sorted value: $0.5 \times 110 + 0.5 \times 113 = 111.5 \text{ days}$

The inter quartile range = Upper Quartile - Lower Quartile

$IQR = Q3-Q1=111.5-100=11.5 \text{ days}$

So 50% of pigs reach bacon weight within a range of 11.5 days.

The following is an example of calculations in R - note that we use the type 6 calculation and the default in R is type 7 which we generally use.

```{r}
y <- c(98, 100, 100, 103, 105, 107, 110, 113, 115)

quantile(y, 0.75, type = 6) - quantile(y, 0.25, type = 6) ## Type 6
quantile(y, 0.75) - quantile(y, 0.25) ## Default which we will generally use going forward
```


**PERCENTILES**

The 5th and 95th percentiles cut off 5% of the most extreme values in the distribution of values for the sample. The 1st and 99th percentiles may be similarly defined. As with quartiles, a list of percentiles may be far more informative than the standard deviation for summarizing the spread
of values about the mean or median, especially when the spread is asymmetrical.

** VARIANCE**

In general, we have $n$ observations, so the general formula for the sample variance is

$s^2 = \frac{1}{n-1}\sum_{i=1}^{n} (y_i - \bar{y})^2$

The units for the variance are always the units of the original measurement squared. If units of measurement were kg (e.g. body weight), then the variance would have units $kg^2$.


**STANDARD DEVIATION**

To have a measure of variability with the same units as the original measurement, we take the square root of the variance. This is the standard deviation of the observations (usual symbol is s).

Sample standard deviation,

$s = \sqrt{\frac{1}{n-1}\sum_{i=1}^{n} (y_i - \bar{y})^2}$


**COEFFICIENT OF VARIATION**

The coefficient of variation (CV) is used is used to aide in comparing the variability of two samples that have widely differing means. It is usually expressed as a percentage, and has no units.

$CV = \frac{s}{\bar{y}} \times 100\%$


# 1.2 Graphical Summaries

## 1.2.1 Bar Graphs and Histograms

Frequency tabulations can be represented as a graph of frequency (raw or percentage) against the measurement variable. Discrete data are sometimes expressed as a bar graph where bars are spaced equidistantly along the horizontal axis. Figure 1.2 is a relative frequency histogram
(also known as a percentage frequency histogram) of the sedge plant data. The data represents data from sedge plant counts in 800 1m x 1m quadrats Ideally this data set (since it is discrete) would be plotted as a bar graph.


```{r, fig.cap ="Figure 1.2 Relative frequency histogram of sedge plant data."}
require(tidyverse)
require(ggplot2)

sedge <- read_csv("data/Sedge.csv", show_col_types = FALSE)

ggplot(sedge, aes(x = Plants)) +
  geom_histogram(aes(y = after_stat(count)/sum(after_stat(count)) * 100), bins = 8) +
  ylab("Percentage") +
  xlab("Number of plants per quadrat")

```

Continuous measurements are free to take any whole or fractional number within their range e.g. plant height, soil pH, concentration of nitrates in a water sample. Histograms (with bars touching each other) are the norm for continuous data.

```{r, fig.cap ="Figure 1.3 Frequency histogram of creeping bentgrass data."}

bentgrass <- read_csv("data/Bentgrass.csv", show_col_types = FALSE)

ggplot(bentgrass, aes(x = Root_length_mm)) +
  geom_histogram(bins = 20) +
  xlab("Root length (mm)")

```

A histogram can give nearly complete information about the distribution of data. For example, from Figure 1.3 above (that shows a fairly symmetric distribution) we can estimate that the mean ≈ 95 mm (the centre of the data) and standard deviation ≈ 15 mm (since for symmetric distributions approximately 95% of data lies within 2 standard deviations either side of the mean i.e. a total of 4 standard deviations across 95% of the data values).

A histogram needs a relatively large sample size for it to be informative (i.e. 30 or more data values).


## 1.2.2 Boxplots

Boxplots show the shape of the distribution of data very clearly and are also helpful in identifying any outlying (or extreme) values.

Example 1 Creeping bentgrass turf was laid in an experiment to assess root growth. Eighty (80) “plugs” were randomly sampled 4 weeks after laying. Root growth was measured by averaging the length (mm) of the ten longest roots in each plug.

```{r}
ggplot(bentgrass, aes(x = Root_length_mm)) +
  geom_boxplot() +
  xlab("Root length (mm)")
```


Notes on boxplots:

- 50% of the data are contained within the box (inter quartile range).
- Whiskers are extended to a maximum of 1.5 x IQR 
- Any data values beyond these maximum whisker lengths are plotted individually, usually by an asterisk or dot $\Rightarrow$ these may be **outliers** and distort the results of any further analysis
- A boxplot gives useful summary of the shape of the data distribution e.g. Is it symmetric or skewed? Are there any outliers?
- Boxplots do not need as many data values (as some other graphs such as histograms and dot plots) for them to be informative.

**GUIDELINES FOR MAKING A BOXPLOT**

We will use the gravimetric water content of soil (%) from Method A in the irrigation data set to illustrate how a boxplot is constructed. In this example we have n = 10 observations. We can use the R command summary to calculate some of the important values in a boxplot such as Q1, Q2, and Median.

```{r}
soil_water <- data.frame(water_content = c(7.5, 9.0, 9.3, 10.4, 10.4, 10.6, 10.7, 11.6, 12.1, 12.8))

summary(soil_water$water_content)
```


Step 1: Determining the Box: 

- The quartiles (Q1: 1st Qu. & Q3: 3rd Qu.) then become the basis for the “box” 
- The median is shown as a vertical line through the box.

Step 2: Determining (potential) outliers

- Calculate IQR = Inter-quartile range = 11.375 - 9.575 = 1.8
- Calculate Q1 - 1.5 x IQR = 9.225 - 1.5 x 1.8 = 6.875 Since there are no observations smaller than 6.875, there are no low-valued outliers.
- Calculate Q3 + 1.5 x IQR = 11.375 + 1.5 x 1.8 = 14.075 Since there are no observations greater than 14.075, there are no high-valued outliers.
- If any values were flagged to be (potential) outliers, they are plotted as individual points on the boxplot, usually as a "*" or $\bullet$.

Step 3: Determining the “whisker” lengths

- Extend the whisker from the lower end of the box (Q1) to the smallest value that is not an outlier, i.e. to the smallest value greater then Q1 - 1.5 x IQR. That is, the whisker here will be extended down to the minimum value which is 7.5.
- Extend the whisker from the upper end of the box (Q3) to the largest value that is not an outlier, i.e. to the greatest value smaller then Q3 + 1.5 x IQR. That is, the whisker here will be extended up to 12.8.

Using these values, the following boxplot is obtained:

```{r}
ggplot(soil_water, aes(x = water_content)) +
  geom_boxplot() +
  xlab("gravimetric water content of soil (%)")
```



# 1.3 Shapes of Distributions

## 1.3.1 Symmetric

This is a distribution such that the left hand side of the frequency polygon is a mirror image of the right hand side. For symmetrical distributions, the mean, median and mode all have the same value. Substantial differences in these three statistics could provide valuable information about the data set (as we’ll see in the sections on positively and negatively skewed distributions). Some examples of symmetric distributions appear below.
5th
6th
10.4
10.6
0.5
2nd
3rd
9.0
9.3
0.75
8th
9th
11.6
12.1
0.25
10
Figure 1.4 Four examples of symmetric distributions.
Example: Creeping bentgrass: root growth (mm)
Summary statistics for Root
Number of observations = 80
Number of missing values = 0
Mean = 93.86
Median = 93.00
Minimum = 65.00
Maximum = 135.00
Range = 70.00
Lower quartile = 86.50
Upper quartile = 102.00
Standard deviation = 13.56
Variance = 183.99
Skewness = 0.08
Kurtosis = 0.35
Note that the mean and the median are similar (93.9, and 93.0 mm respectively). This is indicative of symmetric data.
The boxplot and histogram below show that the data is symmetric about the mean. You will notice from the boxplot that there is one high value outlier indicated by the “x” at the right hand side of the graph. The “4” indicates that this is the 4th observation in the data set i.e. the value 135mm.
11
Figure 1.5 Boxplot (left) and histogram (right) of the creeping bentgrass data show symmetry.
1.3.2 Positively Skewed
For right-skewed distributions we find that the mode (if one exists) is always less than the
median and the median is always less than the mean.
Example:
As part of an evaluation of a clean-up of a contaminated site, 100 soil samples were taken
randomly across an area and the level of 1,2,3,4 Tetrachlorobenzene was recorded in parts per
billion (TcCB, ppb). The following descriptive analysis was undertaken.
Summary statistics for TcCB
Number of observations = 100
Number of missing values = 0
Mean = 1.41
Median = 0.57
Minimum = 0.01
Maximum = 26.60
Range = 26.59
Lower quartile = 0.23
Upper quartile = 1.29
Standard deviation = 3.10
Variance = 9.60
Skewness = 6.00
Kurtosis = 43.12
4
80 90 100 110 120 130
Root
70
Histogram for Root
70
20
80
0
100
5
120
10
15
90 110 130
25
12
Figure 1.6 Boxplot (left) and histogram (right) of the TcCB data showing positive skewness.
The distribution is highly positively skewed (right skewed): there are extreme outliers at high
levels. This is also demonstrated by the mean (1.412 ppb) being substantially greater than the
median (0.570 ppb).
1.3.3 Negatively Skewed
For left-skewed distributions we find that the mode is greater than the median and the median is
greater than the mean.
Example:
Data from 50 river pollution monitoring sites were reviewed over a 28 day period, and the
number of days (out of 28) that the pollution standard had been complied with was recorded.
Summary statistics for Days
Number of observations = 50
Number of missing values = 0
Mean = 25.18
Median = 26.00
Minimum = 19.00
Maximum = 28.00
Range = 9.00
Lower quartile = 24.00
Upper quartile = 26.00
Standard deviation = 2.00
Variance = 3.99
Skewness = -1.20
Kurtosis = 1.28
Although it appears that the median has not been drawn into the boxplot below, it is just that the
median is very close to Q3 (the right hand side of the box) and the poor resolution in the
graphics has not been able to distinguish the two.
99
8084
98
92
64
27
94
79
881042
0 5 10 15 20 25
TcCB
Histogram for TcCB
0 10 15 20 30
80
60
40
20
25
0
5
13
Figure 1.7 Boxplot (left) and histogram (right) of the river pollution data showing negative
skewness.
The distribution is negatively skewed (left skewed): there are some outliers at low levels. This
is also demonstrated by the mean (25.2 days) being marginally less than the median (26 days),
although the effect is not very strong.
1.3.4 Statistics of Shape
There are two statistics useful for describing shape.
 SKEWNESS
Skewness is another name for asymmetry which means that one tail of the frequency
distribution is drawn out more than the other. A skewness of zero implies a symmetrically
shaped histogram, a negative value implies skewness to the left and a positive value implies
skewness to the right.
 KURTOSIS
Kurtosis is a measure of how “peaked” (leptokurtic) a frequency distribution is or how “flattened”
(platykurtic) it is. A negative value indicates platykurtosis (or flatness), and a positive value
indicates leptokurtosis (peakedness).
