---
title: Lab 02 -- Statistical programming
embed-resources: true
resources:
  - data/soil.xlsx
---

```{r setup}
#| include: false
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, readxl, modeest)
```

::: {.callout-tip}
## Learning Outcomes

At the end of this practical students should be able to:

- Acknowledge the importance of academic integrity
- Import, subset and inspect data in R
- Calculate simple summary statistics using both R and Excel
- Generate dynamic reports in Quarto using inline R code
- understand how to debug R code
:::

## Before we begin

At the beginnning of each lab, please create a new Quarto document in your project folder. This will be where you will practice R code, write notes and work on the exercises in this lab.

The following packages are required for this practical:

- `pacman` -- a package for managing the loading of other packages (among other awesome features)
-`tidyverse` -- a collection of R packages designed for data science
- `readxl` -- a package for reading Excel files
- `modeest` -- a package for estimating the mode of a distribution

If you have not already installed these packages, you can do so by adding the following code into a code chunk in your Quarto document:

```{r}
#| eval: false
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, readxl, modeest)
```

Finally, please download the data file used in this lab here: [soil.xlsx](data/soil.xlsx).

## Academic integrity

This exercise encourages students to discuss academic integrity, and in particular the grey areas often present. Your demonstrator will provide you with a number of scenarios to discuss with each other in smaller groups, and then with the class. 

If you are interested in more information on Academic Integrity at the University of Sydney, see the following link: [Academic Integrity](https://www.sydney.edu.au/students/academic-integrity.html). Also ensure you have completed the Academic Honesty Education Module (AHEM). This must be complete before your first assessment is due (next week for ENVX1002). 

## Exercise 1: Setting up

Today we will go through the process of getting into your RStudio project for labs and set up a new Quarto document. Follow these steps carefully:

1. **Open RStudio Project**
   - Locate your RStudio project file (`.Rproj`) in your lab folder
   - Either:
     a. Open RStudio and use File > Open Project to navigate to your `.Rproj` file
     b. Double-click the `.Rproj` file directly from your file explorer

::: {.column-margin}
Look at the top-right corner of your RStudio window. If you've opened the project correctly, you'll see your project name displayed there. This confirms you're working in the right project environment.
:::

2. **Create a New Quarto Document**
   - In RStudio, click File > New File > Quarto Document...
   - In the dialog box that appears:
     - Choose "HTML" as the output format
     - Give your document a clear title (e.g., "Lab 02")
     - Click "Create"

3. **Save Your Document**
   - Immediately save your new document
   - Click File > Save or press Ctrl+S (Windows) / Cmd+S (Mac)
   - Name it something meaningful like `lab02_practice.qmd`
   - Ensure you save it in your lab project folder

4. **Render the Document**
   - Click the "Render" button in the editor toolbar
   - Review the output HTML file that appears
   - Let us know if you encounter any issues

::: {.callout-tip}
## Pro Tips
- Look for your project name in the top-right corner to confirm you're in the right project
- Use Ctrl+S (Windows) / Cmd+S (Mac) to save your work frequently
- Check for missing backticks or brackets if your document won't render
:::

## Exercise 2: water quality

## Exercise 3: summary statistics

## Exercise 4: interpreting spread

## Exercise 5: debugging issues